---
title: "Fitbit Activity Analysis"
author: "BT"
date: "2026-03-05"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    highlight: tango
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r Libraries}
library(tidyverse)
library(lubridate)
library(plotly)
library(knitr)
```
Libraries


```{r}
master_df <- read_csv("fitbit_master_database.csv", guess_max = 100000)

analysis_df <- master_df %>%
  filter(minute >= ymd_hms("2026-01-01 00:00:00"))
```
Import & Prepare Data
<br><br>

### **1. Step Count Comparison**
This visualization highlights the "Step Gap"—the difference between the base daily movement and the movement generated during tracked gym sessions.

```{r Steps Comparison}
ambient_steps <- analysis_df %>%
  filter(Workout_Type == "None") %>%
  mutate(date = as_date(minute)) %>%
  group_by(date) %>%
  summarize(steps = sum(steps, na.rm = TRUE)) %>%
  mutate(Type = "Ambient (Non-Gym)")

total_steps <- analysis_df %>%
  mutate(date = as_date(minute)) %>%
  group_by(date) %>%
  summarize(steps = sum(steps, na.rm = TRUE)) %>%
  mutate(Type = "Total Daily Steps")

step_comparison <- bind_rows(ambient_steps, total_steps)

p1 <- ggplot(step_comparison, aes(x = date, y = steps, color = Type, group = Type)) +
  geom_line(linewidth = 0.5) +
  geom_point(size = 1) +
  
  geom_text(
    data = filter(step_comparison, Type == "Total Daily Steps" & steps > 10000),
    aes(label = paste0(format(date, "%b %d"), "\n", format(steps, big.mark = ","))),
    nudge_x = 2.6,
    fontface = "bold",
    size = 3,
    show.legend = FALSE
  ) +
  
  scale_color_manual(values = c("Ambient (Non-Gym)" = "#9467bd", "Total Daily Steps" = "#2ca02c")) +
  theme_minimal() +
  labs(title = "Total vs. Ambient Daily Step Count",
       subtitle = "Markers indicate days exceeding 10,000 total steps",
       x = "Date", y = "Steps", color = "Measurement") +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.2)), labels = scales::comma) +
  theme(legend.position = "right")

ggplotly(p1)
```

**Analysis:** The green line represents your full daily achievement, while the purple line shows the baseline "lifestyle" movement. The gaps between these lines indicate days where gym sessions (Cardio or Warm-ups) contributed significantly to the overall. Notice how on high-step days, the ambient movement often stays consistent, suggesting the gym is an addition to activity rather than a replacement.

### **2. Heart Rate Density**
Understanding which zones spend the most time in helps categorize the "strain" of each workout modality.

```{r Heart Rate Density}
workout_df <- analysis_df %>%
  filter(Workout_Type != "None") %>%
  filter(!is.na(heart_rate))

p2 <- ggplot(workout_df, aes(x = heart_rate, fill = Workout_Type)) +
  geom_density(alpha = 0.6, color = "white", linewidth = 0.5) +
  scale_fill_manual(values = c("Warm up" = "#f28e2b", "Cardio" = "#e15759", "Weightlifting" = "#4e79a7")) +
  theme_minimal() +
  labs(title = "Heart Rate Distribution by Modality",
       x = "Heart Rate (BPM)", y = "Density")

ggplotly(p2)
```

**Analysis:** Cardio (red) shows a distinct right-shift, spending the majority of time in high-aerobic zones. Weightlifting (blue) displays a broader, multi-modal distribution; this reflects the intermittent nature of lifting, where heart rate spikes during a set and recovers during rest periods.

### **3. Burn Efficiency**
We evaluate the relationship between cardiovascular strain and caloric expenditure to determine which sessions are most "efficient."

```{r HR vs Calories Scatter}
p3 <- ggplot(workout_df, aes(x = heart_rate, y = calories, color = Workout_Type)) +
  geom_point(alpha = 0.4, size = 1.5) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, linewidth = 1) +
  scale_color_manual(values = c("Warm up" = "#f28e2b", "Cardio" = "#e15759", "Weightlifting" = "#4e79a7")) +
  theme_minimal() +
  labs(title = "Caloric Burn vs. Cardiovascular Exertion",
       x = "Heart Rate (BPM)", y = "Calories/Min")

ggplotly(p3)
```

**Analysis:** This scatterplot shows a strong linear correlation across all types. However, the slopes of the trend lines reveal that Weightlifting often generates a higher caloric "return" at lower heart rates compared to steady-state Cardio, likely due to the metabolic cost of moving heavy external loads.

<br>
Looking into the outlier:

```{r Max Outlier}
workout_df %>%
  slice_max(calories, n = 1) %>%
  select(minute, Workout_Type, heart_rate, calories)
```

```{r Outlier Investigation}
# 1. Capture the exact timestamp of the highest calorie minute
outlier_row <- workout_df %>%
  slice_max(calories, n = 1)

outlier_time <- outlier_row %>% pull(minute)

# 2. Find the start time of the specific WEIGHTLIFTING block on that day
lift_start <- workout_df %>%
  filter(as_date(minute) == as_date(outlier_time), 
         Workout_Type == "Weightlifting") %>%
  summarize(start_time = min(minute)) %>%
  pull(start_time)

# 3. Calculate the difference (Time into the Weightlifting session)
time_offset <- difftime(outlier_time, lift_start, units = "mins")

# Display the results
paste("The outlier occurred at", outlier_time, 
      "which was", time_offset, 
      "minutes into the Weightlifting session.")
```

```{r Monday Weightlifting Review}
# 1. Compare all Monday WEIGHTLIFTING workouts
monday_review <- workout_df %>%
  # Isolation Step: Only look at the modality where the outlier occurred
  filter(Workout_Type == "Weightlifting") %>%
  # Filter for Mondays
  filter(wday(minute, label = TRUE) == "Mon") %>%
  # Group by the specific date
  mutate(date = as_date(minute)) %>%
  group_by(date) %>%
  # Summarize the core stats for the entire session
  summarize(
    total_duration_mins = n(),
    total_calories = sum(calories, na.rm = TRUE),
    avg_heart_rate = round(mean(heart_rate, na.rm = TRUE), 1),
    max_heart_rate = max(heart_rate, na.rm = TRUE)
  )

# 2. Render the table in the portfolio
kable(monday_review, caption = "Monday Weightlifting Performance Comparison")
```

### **4. Monday Weightlifting Deep-Dive**
A comparative review of performance across all Monday sessions, which historically serve as the "volume lead" for the training week.

```{r Monday Weightlifting Comparison Graph}
monday_review <- workout_df %>%
  filter(Workout_Type == "Weightlifting") %>%
  filter(wday(minute, label = TRUE) == "Mon") %>%
  mutate(date = as_date(minute)) %>%
  group_by(date) %>%
  summarize(
    total_duration_mins = n(),
    total_calories = sum(calories, na.rm = TRUE),
    avg_heart_rate = round(mean(heart_rate, na.rm = TRUE), 1)
  )

monday_long <- monday_review %>%
  pivot_longer(cols = c(total_calories, avg_heart_rate), 
                names_to = "metric", values_to = "value") %>%
  mutate(metric = case_when(
    metric == "total_calories" ~ "Total Calories",
    metric == "avg_heart_rate" ~ "Average Heart Rate (BPM)"
  ))

p4 <- ggplot(monday_long, aes(x = date, y = value, fill = metric)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~metric, scales = "free_y") + 
  theme_minimal() +
  scale_fill_manual(values = c("Total Calories" = "#e15759", "Average Heart Rate (BPM)" = "#4e79a7")) +
  labs(title = "Monday Weightlifting: Intensity vs. Volume", x = "Date", y = NULL)

ggplotly(p4)
```

**Analysis:** By faceting Calories against Heart Rate, we can spot the **Feb 23rd** outlier immediately. While Heart Rate stayed within the typical Monday range, the Caloric expenditure spiked significantly, suggesting either a massive increase in lifting volume or a period of sensor instability discussed in the technical audit.

<br><br>
### Final Analysis: Variance of Feb 23rd Performance

The session on **Feb 23rd** exhibited a 60% increase in total caloric expenditure compared to the February Monday average. \* **Temporal Analysis:** A significant outlier (22.7 kcal/min) was identified. By calculating the delta between the Weightlifting start time (10:22 AM) and the peak (10:38 AM), we confirmed the anomaly occurred exactly **16 minutes** into the lifting block. \* **Root Cause:** While the specific lift is unrecorded, the 16-minute mark correlates with a self-reported period of high physical exertion following a 6-day recovery window. \* **Conclusion:** The increased session volume and intensity were genuine results of extended recovery, though the specific magnitude of the peak was likely amplified by sensor instability during heavy mechanical exertion.
